``Interestingness'' in Mathematical Theory Exploration

### 2019-05-14 PhDThesis/background_haskell.tex
Changed Haskell background

We mostly focus our attention on the Haskell programming language, both as an
implementation vehicle and as our representation of functions, properties, etc.
to explore. We choose Haskell since it combines formal, logical underpinnings
which aid reasoning (compared to more popular languages like Java or C), yet it
is still popular enough to sustain a rich ecosystem of tooling and a large
corpus of existing code (compared to more formal languages like Coq or
Isabelle). Haskell is well-suited to programming language research; indeed, this
was a goal of the language's creators \cite{marlow2010haskell}. Like most
\emph{functional} programming languages, Haskell builds upon $\lambda$-calculus,
with extra features such as a strong type system and ``syntactic sugar'' to
improve readability.

The following features make Haskell especially useful for our purposes, although
many are also present in other languages such as StandardML and Coq (which we
also use, but only when needed for compatibility with prior work):

@@

\item{Functional}: All control flow in Haskell is performed by function
  abstraction and application, which we can reason about using standard rules of
  inference such as \emph{modus ponens}.

\item{Pure}: Execution of actions (e.g. reading files) is separate to evaluation
  of expressions; hence our reasoning can safely ignore complicated external and
  non-local interactions.

\item{Statically Typed}: Expressions are constrained by \emph{types}, which can
  be used to eliminate unwanted combinations of values, and hence reduce search
  spaces; \emph{static} types can be deduced syntactically, without having to
  execute the code.

\item{Non-strict}: If an evaluation strategy exists for $\beta$-normalising an
  expression (i.e. performing function calls) without diverging, then a
  non-strict evaluation strategy will not diverge when evaluating that
  expression. This is rather technical, but in simple terms it allows us to
  reason effectively about a Turing-complete language, where evaluation may not
  terminate. For example, when reasoning about \emph{pairs} of values \hs{(x,
    y)} and projection functions \hs{fst} and \hs{snd}, we might want to use an
  ``obvious'' rule such as
  $\forall \text{\hs{x y}}, \text{\hs{x}} = \text{\hs{fst (x, y)}}$. Haskell's
  non-strict semantics makes this equation valid; whilst it would \emph{not} be
  valid in the strict setting common to most other languages, where the
  expression \hs{fst (x, y)} will diverge if \hs{y} diverges (and hence alter
  the semantics, if \hs{x} doesn't diverge).

\item{Algebraic Data Types}: These provide a rich grammar for building up
  user-defined data representations, and an inverse mechanism to inspect these
  data by \emph{pattern-matching}. For our purposes, the useful consequences of
  ADTs and pattern-matching include their amenability for inductive proofs and
  the fact they are \emph{closed}; i.e. an ADT's declaration specifies all of
  the normal forms for that type. This makes exhaustive case analysis trivial,
  which would be impossible for \emph{open} types (for example, consider classes
  in an object oriented language, where new subclasses can be introduced at any
  time).

\item{Parametricity}: This allows Haskell \emph{values} to be parameterised over
  \emph{type-level} objects; provided those objects are never inspected. This
  has the \emph{practical} benefit of enabling \emph{polymorphism}: for example,
  we can write a polymorphic identity function \hs{id :: forall t. t ->
    t}. \footnote{Read ``\hs{a :: b}'' as ``\hs{a} has type \hs{b}'' and ``\hs{a
      -> b}'' as ``the type of functions from \hs{a} to \hs{b}''.} Conceptually,
  this function takes \emph{two} parameters: a type \hs{t} \emph{and} a value of
  type \hs{t}; yet only the latter is available in the function body,
  e.g. \hs{id x = x}. This inability to inspect type-level arguments gives us
  the \emph{theoretical} benefit of being able to characterise the behaviour of
  polymorphic functions from their type alone, a technique known as
  \emph{theorems for free} \cite{wadler1989theorems}.

\item{Type classes}: Along with their various extensions, type classes are
  interfaces which specify a set of operations over a type (or other type-level
  object, such as a \emph{type constructor}). Many type classes also specify a
  set of \emph{laws} which their operations should obey but, lacking a simple
  mechanism to enforce this, laws are usually considered as documentation. As a
  simple example, we can define a type class \hs{Semigroup} with the following
  operation and associativity law:

\begin{lstlisting}[language=Haskell, xleftmargin=.2\textwidth, xrightmargin=.2\textwidth]
op :: forall t. Semigroup t => t -> t -> t
\end{lstlisting}

$$\forall \text{\hs{x y z}}, \text{\hs{op x (op y z)}} = \text{\hs{op (op x y) z}}$$

@@

\item{Modularity}: Haskell has a module system, where each module may specify an
  \emph{export list} containing the names which should be made available for
  other modules to import. When such a list is given, any expressions \emph{not}
  on the list are considered \emph{private} to that module, and are hence
  inaccessible from elsewhere. This mechanism allows modules to provide more
  guarantees than are available just in their types. For example, a module may
  represent email addresses in the following way:

  \begin{haskell}
    module Email (Email(), at, render) where

@@

    render :: Email -> String
    render (E u h) = u ++ "@" ++ h

    at :: String -> String -> Maybe Email
    at "" h  = Nothing
    at u  "" = Nothing
    at u  h  = Just (E u h)
  \end{haskell}

  This \hs{Email} type can be constructed in two ways: directly, via the \hs{E}
  constructor, which requires both a user part and a host part (both
  \hs{String}s, for simplicity); or indirectly, via the \hs{at} function, which
  will only call \hs{E} if both \hs{String}s are non-empty (a programming
  pattern known as a ``smart constructor''). The export list in the first line
  contains the \hs{at} function, but not the \hs{E} constructor~\footnote{The
    syntax \hs{Email()} means we're exporting the \hs{Email} type, but not any
    of its constructors.}, so any code using this module is only able to create
  \hs{Email} values with non-empty parts.

@@

\newcommand{\CVar}{\texttt{Var}}
\newcommand{\CLit}{\texttt{Lit}}
\newcommand{\CApp}{\texttt{App}}
\newcommand{\CLam}{\texttt{Lam}}
\newcommand{\CLet}{\texttt{Let}}
\newcommand{\CCase}{\texttt{Case}}
\newcommand{\CType}{\texttt{Type}}
\newcommand{\CLocal}{\texttt{Local}}
\newcommand{\CGlobal}{\texttt{Global}}
\newcommand{\CConstructor}{\texttt{Constructor}}
\newcommand{\CLitNum}{\texttt{LitNum}}
\newcommand{\CLitStr}{\texttt{LitStr}}
\newcommand{\CAlt}{\texttt{Alt}}
\newcommand{\CDataAlt}{\texttt{DataAlt}}
\newcommand{\CLitAlt}{\texttt{LitAlt}}
\newcommand{\CDefault}{\texttt{Default}}
\newcommand{\CNonRec}{\texttt{NonRec}}
\newcommand{\CRec}{\texttt{Rec}}
\newcommand{\CBind}{\texttt{Bind}}

@@

\begin{figure}
  \begin{haskell}
    data Nat = Z
             | S Nat

    even :: Nat -> Bool
    even    Z  = True
    even (S n) = odd n
  \end{haskell}
  \caption{A Haskell datatype for Peano numerals with some simple arithmetic
    functions, including mutually-recursive definitions for \hs{odd} and
    \hs{even}. \hs{Bool} is Haskell's built in boolean type, which can be
    regarded as \hs{data Bool = True | False}.}
  \label{fig:haskellexample}
\end{figure}

@@

Figure~\ref{fig:haskellexample} shows some simple Haskell function definitions,
along with a custom datatype for Peano numerals. The translation to Core syntax
is routine, and shown in Figure~\ref{fig:coreexample}. Although the Core is more
verbose, we can see that similar structure in the Haskell definitions gives rise
to similar structure in the Core; for example, the definitions of \hs{odd} and
\hs{even} are identical in both languages, except for the particular identifiers
used. It is this close correspondence which allows us to analyse Core
expressions in place of their more complicated Haskell source.

Note that we exclude representations for type-level entities, including datatype
definitions like that of \hs{Nat}. GHC can represent these, but in this work we
only consider reducible expressions (i.e. value-level bindings of the form
\mbox{\hs{f a b ... = ...}}).

>>>>

\begin{figure}
  \centering
  \begin{haskell}
-- A datatype with two constructors
data Bool = True | False

-- A recursive datatype (S requires a Nat as argument)
data Nat = Z | S Nat

-- A polymorphic (AKA "generic") datatype
data List t = Nil | Cons t (List t)

-- Arithmetic functions

-- Mutually-recursive functions

even :: Nat -> Bool
even    Z  = True
even (S n) = odd n
  \end{haskell}
  \caption{Haskell code, defining datatypes and functions involving them (note
    that \hs{--} introduces a comment). \hs{Bool} is the Booleans, \hs{Nat} is a
    Peano encoding of the natural numbers and \hs{List t} are polymorphic lists
    with elements of type \hs{t}. Functions are called using whitespace, e.g.
    \hs{f x}, and may be defined by pattern-matching (case analysis) on their
    arguments. \hs{A :: B} is an optional type annotation, stating that value
    \hs{A} has type \hs{B}. \hs{A -> B} is the type of functions from \hs{A} to
    \hs{B}.}
  \label{fig:haskellexample}
\end{figure}

We mostly focus our attention on the Haskell programming language, both as an
implementation vehicle and as our representation of functions, properties, etc.
to explore. We choose Haskell since it combines formal, logical underpinnings
which aid reasoning (compared to more popular languages like Java or C), yet it
is still popular enough to sustain a rich ecosystem of tooling and a large
corpus of existing code (compared to more formal languages like Coq or
Isabelle). Haskell is well-suited to programming language research; indeed, this
was a goal of the language's creators \cite{marlow2010haskell}. An example of
Haskell code is given in Figure~\ref{fig:haskellexample}.

Like most \emph{functional} programming languages, Haskell builds upon
$\lambda$-calculus, with extra features such as a strong type system and
``syntactic sugar'' to improve readability. The following features make Haskell
especially useful for our purposes, although many are also present in other
languages such as StandardML and Coq (which we also use, but only when needed
for compatibility with prior work):

@@

\item{Functional}: Most control flow in Haskell (all except
  \emph{pattern-matching}, described below) is performed by function abstraction
  and application , which we can reason about using standard rules of inference
  such as \emph{modus ponens}. For example, reading from external sources (like
  environment variables) is \emph{impure}, so the only way to parameterise a
  value of type \hs{B} with a value of type \hs{A} is using a function of type
  \hs{A -> B}. Conversely, applying a function of type \hs{A -> B} to a value of
  type \hs{A} can only ever produce a value of type \hs{B} (it may instead crash
  the program or loop forever, but neither of those are \emph{observable}, i.e.
  we can't branch on them).

\item{Pure}: Execution of actions (e.g. reading or writing files) is separate to
  evaluation of expressions; hence our reasoning can safely ignore complicated
  external and non-local interactions. Purity ensures \emph{referential
    transparency}: references can be substituted for their referent with
  no change in semantics (as in standard mathematical practice). This implies
  that calling a function twice with the same argument must produce the same
  result (modulo crashes or infinite loops). For example, consider applying the
  function \hs{pair x = (x, x)} to some arbitrary function call \hs{foo y}. The
  references \hs{x} in the resulting pair \hs{(x, x)} both refer to \hs{foo y}
  so, by referential transparency, can be substituted to get
  \hs{(foo y, foo y)}. Both members of \hs{(x, x)} are identical, by definition;
  hence, to preserve the semantics, both (and indeed \emph{all}) calls to
  \hs{foo y} must also be identical. This holds regardless of what we choose for
  \hs{foo} and \hs{y}, and implies that function behaviour cannot depend on
  external interactions or non-deterministic processes (which may change between
  calls).

\item{Statically Typed}: Expressions are constrained by \emph{types}, which can
  be used to eliminate unwanted combinations of values (for example
  \hs{plus True} is not a valid expression), and hence reduce search spaces;
  \emph{static} types can be deduced syntactically, without having to execute
  the code. The type of Haskell expressions can also be \emph{inferred}
  automatically.\footnote{Except for certain cases, such as those caused by the
    \emph{monomorphism restriction}~\cite{marlow2010haskell}[$\S$4.5.5]}

\item{Curried}: All functions in Haskell take a single argument (as in
  $\lambda$-calculus), which makes them easier to manipulate programatically.
  Currying allows multi-argument functions to be simulated, by accepting one
  argument and returning a function to accept the rest. The \hs{mult} function
  in Figure~\ref{fig:haskellexample} has type \hs{Nat -> Nat -> Nat} meaning it
  takes a \hs{Nat} as argument and returns a function of type
  \hs{Nat -> Nat}. Function calls are written with whitespace, so \hs{mult x y}
  calls the \hs{mult} function with the argument \hs{x}, then calls the result
  of that with the argument \hs{y}. This allows \emph{partial application} such
  as \hs{double = mult (S (S Z))}, but for our purposes it is important for
  unifying calling conventions. For example, in Javascript \hs{mult x y} would
  be either \js{mult(x, y)} or \js{mult(x)(y)}, and depends on the definition of
  \js{mult} (the problem compounds with more arguments). In Haskell there is no
  distinction between these forms.

\item{Non-strict}: ``Strict'' evaluation strategies evaluate the arguments of a
  function call before the function body; non-strict does the opposite. The
  Haskell standards do not specify a particular evaluation strategy, but they
  do require that it be non-strict (for efficiency, most implementations use
  \emph{lazy} evaluation to avoid duplicating work). Strictness can result in
  infinite loops and other errors which may be avoided by non-strictness,
  making the latter more useful for reasoning in the face of such errors.
  For example, given a pair of values \hs{(x, y)} and a projection function
  \hs{fst}, we might make the ``obvious''  conjecture that
  \hs{fst (x, y) = x}.\footnote{Incidentally, this is also a valid definition of
    the \hs{fst} function} This statement is true for non-strict languages, but
  \emph{not} for strict languages. Crucially, a strict language will attempt to
  calculate the value of \hs{y}, which may cause an infinite loop or other
  error; a non-strict language like Haskell will ignore \hs{y} since it isn't
  used in the body of the \hs{fst} function.

\item{Algebraic Data Types}: These provide a rich grammar for building up
  user-defined data representations, and an inverse mechanism to inspect these
  data by \emph{pattern-matching} (Haskell's other form of control flow). The
  \hs{Bool}, \hs{Nat} and \hs{List t} definitions in
  Figure~\ref{fig:haskellexample} are ADTs; whilst the functions use
  pattern-matching to branch on their first argument. For our purposes, the
  useful consequences of ADTs and pattern-matching include their amenability for
  inductive proofs and the fact they are \emph{closed}; i.e. an ADT's
  declaration specifies all of the normal forms for that type. This makes
  exhaustive case analysis trivial, which would be impossible for \emph{open}
  types (for example, consider classes in an object oriented language, where new
  subclasses can be introduced at any time).

\item{Parametricity}: This allows Haskell \emph{values} to be parameterised over
  \emph{type-level} objects; provided those objects are never inspected. This
  enables \emph{polymorphism}. The \hs{List t} type in
  Figure~\ref{fig:haskellexample} is an example: there are many useful functions
  involving lists which work in the same way regardless of the element type
  (e.g. getting the length, appending, reversing, etc.). An even simpler example
  is the polymorphic identity function \hs{id x = x}. The type of \hs{id} is
  \hs{forall t. t -> t}\footnote{The \hs{forall t.} is optional; type-level
    identifiers beginning with a lowercase letter are assumed to be universally
    quantified variables.}, which we can view as taking \emph{two} parameters:
  a type \hs{t} and a value of type \hs{t}. Only the latter argument appears in
  the definition (as \hs{x}), meaning that we can't use the type \hs{t} to
  determine the function's behaviour. Indeed, in the case of \hs{id} we can't
  branch on the value of \hs{x} either, since we don't know what type it might
  have (our definition must work \emph{for all} types \hs{t}); the only
  functions we can call on \hs{x} must also be polymorphic, and hence also
  incapable of branching. The type of \hs{id} states that it returns a value of
  type \hs{t}; without knowing what that type is, the only type-correct value it
  can return is the argument \hs{x}. Hence the \emph{type} of \hs{id} tells us
  everything about its behaviour, with this style of reasoning known as
  \emph{theorems for free}~\cite{wadler1989theorems}. Haskell definitions are
  commonly made polymorphic like this, to prevent incorrect implementations
  passing the type checker, e.g. \hs{fst :: (Nat, Nat) -> Nat} might return the
  wrong element, but \hs{fst :: (t1, t2) -> t1} can't.

\item{Type classes}: Along with their various extensions, type classes are
  interfaces which specify a set of operations over a type or other type-level
  object (like a \emph{type constructor}, e.g. \hs{List}). Many type classes
  also specify a set of \emph{laws} which their operations should obey but,
  lacking a simple mechanism to enforce this, laws are usually considered as
  documentation. As a simple example, we can define a type class \hs{Semigroup}
  with the following operation and an associativity law:

  \begin{center}
    \begin{haskell}
op :: forall t. Semigroup t => t -> t -> t
    \end{haskell}

    $\forall \text{\hs{x y z}}, \text{\hs{op x (op y z)}} =
                                \text{\hs{op (op x y) z}}$
  \end{center}

@@

\item{Modularity}: Haskell's module system allows definitions to be kept
  private. This mechanism allows modules to provide more guarantees than are
  available just in their types, by constraining the ways that values can be
  constructed. For example, the following module represents email addresses as a
  pair of \hs{String}s, one for the user part and one for the host:

  \begin{haskell}
-- Exports appear between the parentheses
module Email (Email(), at, render) where

@@

render :: Email -> String
render (E user host) = user ++ "@" ++ host

-- if/then/else is sugar for pattern-matching a Bool
at :: String -> String -> Maybe Email
at user host = if user == "" || host == ""
                  then Nothing
                  else Just (E user host)
  \end{haskell}

  An \hs{Email} value can be constructed by passing any two \hs{String}s to
  \hs{E}, but \hs{E} is private (not exported). The \hs{at} function is
  exported, but only passes its arguments to \hs{E} iff they are not empty.\footnote{\hs{Maybe t} is a
    safer alternative to the \textsc{Null} construct of other
    languages~\cite{hoare2009null}. It is defined as
    \hs{data Maybe t = Nothing | Just t} and can be understood as an optional
    value, or a computation which may fail, or as a list with at most one
    element, or as a degenerate search tree with no
    backtracking~\cite{wadler1985replace}} Since this module never calls \hs{E}
  with empty \hs{String}s, and other modules must use \hs{at}, we're guaranteed
  that \emph{all} \hs{Email} values will have non-empty \hs{String}s. Such
  ``smart constructors'' can guarantee \emph{any} decidable property, at the
  cost of performing run-time checks on each invocation.\footnote{We can
    guarantee non-emptiness ``by construction'', without run-time checks or
    \hs{Maybe} wrappers, by changing the type to require at least one element.
    \hs{String} is equivalent to \hs{List Char}, with \hs{""} represented as
    \hs{Nil}. Changing \hs{Nil} to require a \hs{Char} would eliminate empty
    \hs{String}s, e.g. \hs{data NonEmpty t = Nil t | Cons t (NonEmpty t)}. We
    could also use a pair like \hs{data NonEmpty t = NE t (List t)} instead.
    Such precise types are often more desirable than smart constructors, but are
    less general since ad-hoc representations need to be invented to enforce
    each guarantee.}

@@

@@

@@

GHC's translation from Haskell to Core is routine: Figure~\ref{fig:coreexample}
shows the Core representation of functions from Figure~\ref{fig:haskellexample}.
Although the Core is more verbose, we can see that similar structure in the
Haskell definitions gives rise to similar structure in the Core; for example,
the definitions of \hs{odd} and \hs{even} are identical in both languages,
except for the particular identifiers used. It is this close correspondence
which allows us to analyse Core expressions in place of their more complicated
Haskell source.

Note that we exclude representations for type-level entities, including datatype
definitions like that of \hs{Nat}. GHC can represent these, but we are mostly
concerned with dynamic behaviour, which excludes types (they are \emph{erased}
during compilation). For simplicity we also avoid using constructors directly,
choosing instead to wrap them with normal functions (e.g. \hs{s x = S x}), since
this makes our domain more symmetric and reduces cross-language differences.
